Composition-climate paper

### 2016-08-19 Composition_Ecology.Rmd
Added the figures folder to the `gitignore`.
Text changes up through the Introduction to the paper.

Three new references for the `bib` file.

*The introduction should summarize briefly the background and aims, and end with a very brief statement of what has been achieved by the work.*

Global change ecologists and biogeographers use contemporary distributions of species, and, increasingly, paleohistorical datasets to map distributions of species and species traits within environmental space [@moritz2013future;@maguire2015modeling;@ivory2016niche;@pearman2008niche]. Distributions of species within climate space, combined with future climate scenarios and species distribution models, can be used to assess species exposure and sensitivity to climate change, identify species and habitats of concern, set conservation priorities, and prepare for shifts in habitat suitability caused by climate change  [@pellatt2012using;@iverson2013tree;@iverson1998predicting;@anderson2013nichemodels;@guisan2013predicting]. Forest inventory datasets, such as the US Forest Inventory Analysis (FIA), are particularly rich observational datasets that have, *e.g.* been used to assess early signals of range shifts via the distributions of mature trees and seedlings [@monleon2015evidence;@zhu2012failure] and to parameterize models of current and future tree distributions [@iverson2008estimating;@iverson1998predicting].

Increasingly, ecological models and their predictive capacity are being improved by integrating contemporary observational data with paleohistorical data, when available [@moritz2013future;@maguire2015modeling].  Future projections that rely only on contemporary observational data are limited by a narrow time window that may obscure greater plasticity in species-climate responses [@maiorano2013building], incompletely delimit species fundamental niches that lie outside the current realized environmental space [@jackson2000responses], miss the responses of species to past no-analog climates [@veloz2012no], and fail to detect the effects of historical legacies and contingency on species distributions [@svenning2015influence]. Paleoecological records from the last deglaciation show that the time lag of forest responses to abrupt climate change varies from a few years to centuries [@williamsinpress;@webb1986vegetation;@ordonez2013realized;@svenning2015influence]. Habitat loss through human agency can impact apparent species distributions at timescales of decades [@early2014climatic], thousands [@sarmento2013impacts] to hundreds of thousands of years [@ivory2016niche]. In principle, dynamic vegetation models [*e.g.*, @leiblein2016modelling] can represent the physiological, demographic, and dispersal processes that cause tree populations to lage climate change.  In practice, however, parameterization of these models is critical, and processes operating at large temporal and spatial scales remain poorly constrained [@moorcroft2006close].

Over the last several centuries, climate change and land use have both substantially affected species distributions and vegetation structure [*e.g.*, @gehrig2007tree]. Understanding the interacting effects of land-use change and climatic change on species-climate relationships is of particular importance, because anthropogenic land use is highly structured and differentially occupies particular habitat types and portions of climate space.  Hence, species distributions and protected natural areas often represent a biased selection from the available environmental space [@scott2001nature].  This biasing can confound studies that use contemporary distributional data to predict species responses to future climate scenarios [@pyke2004habitat].

The interactions between climate and land-use change can be classified into three types: (1) compounding, (2) confounding and (3) counteracting [modified from @pyke2004habitat]. Compounding effects occur when climatic and land-use change cause species distributions to shift in the same direction.  For example, for butterfly species in the northern Sierra Nevada Mountains of California, warming has been compounded by land-use conversion at the lowest elevations, both driving species upslope [@forister2010compounded]. Confounding effects may cause little overall change on species distributions, with land-use pressure falling within the central distribution for the environmental gradient.  The overall spatial and climatic distribution of Garry oak (*Quercus garryana*) in the Pacific Northwest has changed little over the past several centuries, but land-use conversion has excluded it from the deep-soil sites it formerly occupied [@macdougall2004defining;@pellatt2014environmental].  Hence, predictive models for Garry oak [@pellatt2012using] using only contemporary data may be confounded by this range-wide interaction between human land use and soil characteristics. Counteracting effects occur when climate and land use change act in opposite directions, narrowing the available geographic and climatic space available for species. For example, bird species in the Sierra Nevadas should show downslope migration in response to shifts in precipitation [@tingley2012push], but this downslope migration is limited by land-use conversion in the lowlands.

@@

```
![figone](figures/Fig_1_Composite.png)

@@

In the upper Midwestern US, tree distributions and community composition over the last several centuries have been heavily influenced both by the land uses associated with EuroAmerican Settlement in the 1800s [@gorng2016preset;@schulte2007homogenization;hanberry2012comparison;@rhemtulla2009historical] and by the climate changes accompanying the end of the Little Ice Age [@hotchkiss2007response;@paquette2013climatic] and current anthropogenic warming [@wicci2011climate].  Modern forests, as represented by the Forest and Inventory Analysis [@woudenberg2010forest], show greater homogeneity, weaker ecotones, and significantly different species associations than forests surveyed as part of the Public Land Survey System, a survey that covered the western United States prior to widespread land-use conversion in the late 1800s [@gorng2016preset;@schulte2007homogenization]. There has been widespread conversion of wildland ecosystems to agricultural and urban land use [@rhemtulla2009historical], particularly along the historical prairie and savanna margins of the region (`r fig_nums("land_use_clim", display = "cite")`f).  Forested regions in the northern Great Lakes States have been also affected by widespread timbering during the late 19th and early 20th-century and ongoing land management REF.  In northern forested regions, we may expect to see expanded ranges of early successional species and reduced abundances of late-successional species, following widespread logging.  Additionally, some tree species, *Ulmus americana* and *Castanea dentata* have experienced rangewide population collapses caused by the introduction of exotic pathogens REFS. These land use changes are superimposed on strong regional gradients of temperature and precipitation in the upper Midwest and two major ecotones, one from prairie/savanna to closed forest, in an approximately west to east direction, and one from conifer dominated, sub-boreal forests to deciduous forest in a north to south direction.

Here we show that the climatic distributions of tree genera in the upper Midwest have significantly shifted over the last 100 years as a result of historic climate change, land use, and pathogen outbreaks, and we assess the relative impacts of regional climate shifts and vegetation change in driving these changes.  To this end, we use gridded historic and contemporary climate observations (PRISM) and estimates of pre-settlement and modern forest vegetation [@gorng2016preset] to map the geographic and climatic distributions for 15 major tree taxa in the upper Midwestern United States.  We construct four observed and hypothetical sets of tree-climate relationships, using a 2x2 factorial of pre-settlement and modern vegetation and climates and develop a new method, based on Hellinger distances among tree distributions, to quantify the total change and relative importance of historic vegetation and climate change. We assess whether climate and vegetation changes are compounding, counteracting or confounding based on the magnitude and direction of species-climate shifts. We use this framework to discuss the potential impacts of land-use conversion on species resilience to climate change as predicted by species-distribution models generated from modern distributional data.

>>>>

Global change ecologists and biogeographers use contemporary distributions of species, and, increasingly, paleohistorical datasets to map distributions of species and species traits within environmental space, study underlying processes, and prepare for climate change  [@lamanna2014functional;@moritz2013future;@maguire2015modeling;@ivory2016niche;@pearman2008niche]. Distributions of species within climate space, combined with future climate scenarios and species distribution models, can be used to assess species exposure and sensitivity to climate change, identify species and habitats of concern, set conservation priorities, and prepare for shifts in habitat suitability caused by climate change  [@pellatt2012using;@iverson2013tree;@iverson1998predicting;@anderson2013nichemodels;@guisan2013predicting]. Forest inventory datasets, such as the US Forest Inventory Analysis (FIA), are particularly rich observational datasets that have, been used, for example, to assess early signals of range shifts via the distributions of mature trees and seedlings [@monleon2015evidence;@zhu2012failure] and to parameterize models of current and future tree distributions [@iverson2008estimating;@iverson1998predicting].

Increasingly, ecological models and their predictive capacity are being improved by integrating contemporary observational data with paleohistorical data, when available [@moritz2013future;@maguire2015modeling].  Future projections that rely only on contemporary observational data are limited by a narrow time window that may obscure plasticity in species-climate responses [@maiorano2013building], incompletely delimit species fundamental niches that lie outside the current realized environmental space [@jackson2000responses], miss the responses of species to past no-analog climates [@veloz2012no], and fail to detect the effects of anthropogenic land use, and other historical legacies on species distributions [@svenning2015influence]. Paleoecological records from the last deglaciation show that the time lag of forest responses to abrupt climate change varies from a few years to centuries [@williamsinpress;@webb1986vegetation;@ordonez2013realized;@svenning2015influence], suggesting that tree distributions have yet to fully respond to recent climate changes. In principle, dynamic vegetation models [*e.g.*, @leiblein2016modelling] can represent the physiological, demographic, and dispersal processes that cause tree populations to lage climate change.  In practice, however, parameterization of these models is critical, and processes operating at large temporal and spatial scales remain poorly constrained [@moorcroft2006close]. In contrast t the several studies examining the stability of the climatic niche at timescales of millenia and longer, the effect of 19^th^ and 20^th^ century climate change and land use on species-climate relationships remains largely unknown and understudied.

Over the last several centuries, climate change and land use have both substantially affected species distributions and vegetation structure [*e.g.*, @gehrig2007tree] and may cause species distributions to incompletely fill available climates, or otherwise be at disequilibrium [@early2014climatic; @sarmento2013impacts; @ivory2016niche].  Anthropogenic land use is highly structured and differentially occupies particular habitat types and portions of climate space.  Hence, species distributions and protected natural areas often represent a biased selection from the available environmental space [@scott2001nature].  This biasing can confound studies that use contemporary distributional data to understand the environmental controls on species distributions and predict species responses to future climate scenarios [@pyke2004habitat].

The interactions between climate and land-use change can be classified into three types: (1) compounding, (2) confounding and (3) counteracting [modified from @pyke2004habitat]. Compounding effects occur when climatic and land-use change cause species distributions to shift in the same direction.  For example, in the northern Sierra Nevada Mountains of California, warming has been compounded by land-use conversion at the lowest elevations, both driving butterfly species upslope [@forister2010compounded]. Confounding effects may cause little overall change on species distributions, with land-use pressure falling within the central distribution for the environmental gradient.  The overall spatial and climatic distribution of Garry oak (*Quercus garryana*) in the Pacific Northwest has changed little over the past several centuries, but land-use conversion has excluded it from the deep-soil sites it formerly occupied [@macdougall2004defining;@pellatt2014environmental].  Hence, predictive models for Garry oak [@pellatt2012using] using only contemporary data may be confounded by this range-wide interaction between human land use and soil characteristics. Counteracting effects occur when climate and land use change act in opposite directions, narrowing the available geographic and climatic space available for species. For example, bird species in the Sierra Nevadas are expected to be pulled downslope by increasing precipitation [@tingley2012push], but this downslope migration is limited by land-use conversion in the lowlands.

@@

```
![](figures/Fig_1_Composite.png)

@@

In the upper Midwestern US, tree distributions and community composition over the last several centuries have been heavily influenced by the land uses associated with EuroAmerican Settlement in the 1800s [@goring2016preset;@schulte2007homogenization;@hanberry2012comparison;@rhemtulla2009historical] outbreaks of exotic pathogens [@anagnostakis1987chestnut;@barnes1979stem], and the climate changes accompanying the end of the Little Ice Age [@hotchkiss2007response;@paquette2013climatic] and current anthropogenic warming [@wicci2011climate].  Modern forests, as represented by the Forest and Inventory Analysis surveys [@woudenberg2010forest], show greater homogeneity, weaker ecotones, and significantly different species associations than forests surveyed as part of the Public Land Survey System, a survey that covered the western United States prior to widespread land-use conversion in the late 1800s [@gorng2016preset;@schulte2007homogenization]. Wildland ecosystems have been widely converted to agricultural and urban land use [@rhemtulla2009historical], particularly in historical prairie and savanna ecosystems (`r fig_nums("land_use_clim", display = "cite")`f).  In northern forested regions, ranges of early successional species have expanded and abundances of late-successional species are reduced due to widespread logging in the 19^th^ and 20^th^ century, and ongoing land management [@schulte2007homogenization;@goring2016preset].  Additionally, formerly dominant tree species, *Ulmus americana* and *Castanea dentata* have experienced rangewide population collapses caused by the introduction of exotic pathogens [@barnes1979stem;@anagnostakis1987chestnut]. These vegetation changes are superimposed on strong regional gradients of temperature and precipitation in the upper Midwest and two major ecotones, one from prairie/savanna to closed forest, in an approximately west to east direction, and one from conifer dominated, sub-boreal forests to deciduous forest in a north to south direction.

Here we demonstrate significant shifts in the climatic distributions of tree genera in the upper Midwest over the last 100 years as a result of historic climate change, land use, and pathogen outbreaks, and we diagnose the relative influence of historic climate and vegetation changes on these shifts.  To this end, we use gridded historic and contemporary climate observations (PRISM) and estimates of pre-settlement and modern forest vegetation [@gorng2016preset] to map the geographic and climatic distributions for 15 major tree taxa in the upper Midwestern United States.  We construct four observed and hypothetical sets of tree-climate relationships, using a 2x2 factorial of pre-settlement and modern vegetation and climates.  We develop a new method, based on Hellinger distances among tree-climate distributions, to quantify the total change and relative importance of historic vegetation and climate change. We assess whether climate and vegetation changes are compounding, counteracting or confounding based on the magnitude and direction of species-climate shifts. We use this framework to discuss the potential impacts of historical land-use and climate change on inferences about species resilience to climate change, as predicted by species-distribution models based soley on modern distributional data.
